---
title: "Modeling and assessing pay disparity"
author: "Dawnena Key, Katrin Jõgi"
date: "2023-10-08"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library('tidyr')
library('ggplot2')
library("ggpubr")
theme_set(
  theme_bw() +
    theme(legend.position = "top")
  )
library(GGally)
# Install pwr package
install.packages("pwr") 
library(pwr)
install.packages("car")
library(car)
```

Load data
```{r regression project data}
# Load data 
library('Sleuth3')
data <- case1202[,-c(2)]
attach(data)
```


#Exploratory graphs
#Scatter plots with starting salary and education and seniority
```{r}

e <- ggplot(data, aes(x=Educ, y=Bsal, color = Sex)) + geom_point() + geom_smooth(method=lm,  linetype="dashed", se = FALSE, color="darkred", fill="blue") + labs(title = "Starting salary and education", x = "Education (years)", y = "Starting salary")  + theme(legend.position = c(0.3,0.9)) 
s <- ggplot(data, aes(x=Senior, y=Bsal, color=Sex)) + geom_point() + geom_smooth(method=lm, linetype="dashed", se = FALSE, color="darkred", fill="blue") + labs(title = "Starting salary and seniority", x = "Seniority (months)", y = "Starting salary")  + theme(legend.position = c(0.7,0.9)) 
f <- ggarrange(e, s, ncol = 2, nrow = 1)
f
```


#Boxplot with starting salary and sex, Histogram with starting salary and sex
```{r}
h <- ggplot(data,  aes(x=Bsal, color=Sex, fill=Sex)  ) + geom_histogram(binwidth = 500) + labs(title = "Starting salary,Sex", x = "Starting salary") + theme(legend.position = c(0.8,0.8)) 

b <- ggplot(case1202, aes(x=Sex, y=Bsal)) + geom_boxplot(outlier.colour="red", outlier.shape=8,outlier.size=4) + coord_flip() + labs(title = "Starting salary, Sex", y = "Starting salary")

b1 <- ggplot(case1202, aes(x=Sex, y=Bsal)) + geom_boxplot(outlier.colour="red", outlier.shape=8,outlier.size=4) + labs(title = "Starting salary, Sex", y = "Starting salary")
f2 <- ggarrange(h, b1, ncol = 2, nrow = 1)
f2
```

#Density plot with starting salary and sex
```{r}
ggplot(case1202, aes(Bsal, fill=Sex)) + 
  geom_density(alpha=.5) + 
  scale_fill_manual(values = c('Yellow','#E69F00')) + 
  theme(legend.position = c(0.9,0.6)) + labs(title = "Density for starting salary and sex", x = "Starting salary")
```


#Qplot with starting salary and sex
```{r}
qplot(sample = Bsal, data = data, shape=Sex, color = Sex) + labs(title = "Distribution of Starting salary, Sex") + theme(legend.position = c(0.3,0.8)) 
```

#Qqplots with starting salary and sex
```{r}
ggqqplot(data$Bsal[data$Sex == "Male"]) 
ggqqplot(data$Bsal[data$Sex == "Female"])
```

#Shapiro test for normality
```{r}
shapiro.test(Bsal[Sex=="Male"]) #not normal
shapiro.test(Bsal[Sex=="Female"]) #normal
```

#Variance - Equal
It can be hard to tell if equal variance is satisfied from a graph so we calculate the statistic, and perhaps more importantly we determine if the difference in variance is due to random chance with a hypothesis test.

```{r}
# Extract male and female salaries
male_salaries <- subset(data, Sex == "Male")$Bsal
female_salaries <- subset(data, Sex == "Female")$Bsal

#whether t-test or welch's test, test variances, if equal, use t.test
# F-test 
var.test(male_salaries, female_salaries)

# F-test using formula syntax
var.test(Bsal ~ Sex, data = data)

#Levene's test 
leveneTest(Bsal ~ as.factor(Sex), data = data) 
```

##TWO-SAMPLE TEST
##Go with two-sample t-test, equal variance 

##Equal Variance, T-test is appropriate, two-sample t-test, wilcox test
```{r tests}
t.test(Bsal)
t.test(Bsal[Sex == 'Female'],Bsal[Sex == 'Male'],var.equal = TRUE)
wilcox.test(Bsal ~ Sex, exact=FALSE, correct=TRUE, 
            alternative="two.sided", conf.int=TRUE)  
```

##Power Analysis
```{r}
# Significance level
alpha = 0.05

# Desired power  
power = 0.8

# Estimate effect size 
mean_diff <- mean(Bsal[Sex == "Male"]) - mean(Bsal[Sex == "Female"])
pooled_sd <- sqrt(((length(Bsal[Sex == "Male"])-1)*sd(Bsal[Sex == "Male"])^2 +  
                  (length(Bsal[Sex == "Female"])-1)*sd(Bsal[Sex == "Female"])^2) /  
                 (length(Bsal[Sex == "Male"])+length(Bsal[Sex == "Female"])-2))
d = mean_diff / pooled_sd

# Compute required sample size
pwr.t.test(d = d, power = power, sig.level = alpha,  
           type = "two.sample")
pwr.t2n.test(d = d, power = 0.8, n2 = 32)
```

## FIT MODEL TO DATA

#FORWARD SELECTION

# Model1A. Linear model with Bsal and Educ
```{r}
Educ.lm <- lm(Bsal ~ Educ, data=data)
summary(Educ.lm)
par(mfrow=c(2,2))
plot(Educ.lm)
```

# Model1B. Linear model with Bsal and Senior
```{r}
Senior.lm <- lm(Bsal ~ Senior, data=data)
summary(Senior.lm)
par(mfrow=c(2,2))
plot(Senior.lm)
```

# Model1C. Linear model with Bsal and Exper
```{r}
Exper.lm <- lm(Bsal ~ Exper, data=data)
summary(Exper.lm)
par(mfrow=c(2,2))
plot(Exper.lm)
```

# Model1D. Linear model with Bsal and Age
```{r}
Age.lm <- lm(Bsal ~ Age, data=data)
summary(Age.lm)
par(mfrow=c(2,2))
plot(Age.lm)
```

# Model2A. Non-linear model with Bsal and Educ
```{r}
Educ.poly.lm <- lm(Bsal ~ poly(Educ, 2), data=data)
summary(Educ.poly.lm)
par(mfrow=c(2,2))
plot(Educ.poly.lm)
```

# Model2B. Non-linear model with Bsal and Senior
```{r}
Senior.poly.lm <- lm(Bsal ~ poly(Senior, 2), data=data)
summary(Senior.poly.lm)
par(mfrow=c(2,2))
plot(Senior.poly.lm)
```

# Model2C. Non-linear model with Bsal and Exper
```{r}
Exper.poly.lm <- lm(Bsal ~ poly(Exper, 2), data=data)
summary(Exper.poly.lm)
par(mfrow=c(2,2))
plot(Exper.poly.lm)
```

# Model3A. Transformed model with Bsal and Educ
```{r}
Educ.log.lm <- lm(log(Bsal) ~ log(Educ), data=data)
summary(Educ.log.lm)
par(mfrow=c(2,2))
plot(Educ.log.lm)
```

# Model3B. Transformed model with Bsal and Senior
```{r}
Senior.log.lm <- lm(log(Bsal) ~ log(Senior), data=data)
summary(Senior.log.lm)
par(mfrow=c(2,2))
plot(Senior.log.lm)
```

# Model3C. Transformed model with Bsal and Exper
```{r}
Exper.log.lm <- lm(log(Bsal) ~ Exper, data=data)
summary(Exper.log.lm)
par(mfrow=c(2,2))
plot(Exper.log.lm)
```

# Model4A. Two-variable model with Bsal and Educ + Senior
```{r}
Educ.Senior.lm <- lm(Bsal ~ Educ + Senior, data=data)
summary(Educ.Senior.lm)
par(mfrow=c(2,2))
plot(Educ.Senior.lm)
```

# Model4B. Two-variable non-linear model with Bsal and poly(Educ, 2) + Senior
```{r}
PEduc.Senior.lm <- lm(Bsal ~ poly(Educ, 2) + Senior, data=data)
summary(PEduc.Senior.lm)
par(mfrow=c(2,2))
plot(PEduc.Senior.lm)
```

# Model4C. Two-variable model with Bsal and Educ + Exper
```{r}
Educ.Exper.lm <- lm(Bsal ~ Educ + Exper, data=data)
summary(Educ.Exper.lm)
par(mfrow=c(2,2))
plot(Educ.Exper.lm)
```

# Model5A. Interaction model with Bsal and Educ:Senior
```{r}
ggplot(data, aes(x=Educ, y=Senior, color = Sex)) + geom_point() + geom_smooth(method=lm,  linetype="dashed", se = FALSE, color="darkred", fill="blue") + labs(title = "Seniority and education", x = "Education (years)", y = "Seniority")
Inter.Educ.Senior.lm <- lm(Bsal ~ Educ + Senior + Educ:Senior, data=data)
summary(Inter.Educ.Senior.lm)
par(mfrow=c(2,2))
plot(Inter.Educ.Senior.lm)
```

# Model5B. Interaction model with Bsal and Educ:Exper
```{r}
ggplot(data, aes(x=Educ, y=Exper, color = Sex)) + geom_point() + geom_smooth(method=lm,  linetype="dashed", se = FALSE, color="darkred", fill="blue") + labs(title = "Experience and education", x = "Education (years)", y = "Experience")
Inter.Educ.Exper.lm <- lm(Bsal ~ Educ + Exper + Educ:Exper, data=data)
summary(Inter.Educ.Exper.lm)
par(mfrow=c(2,2))
plot(Inter.Educ.Exper.lm)
```

# Model5C. Interaction model with Bsal and Senior:Exper
```{r}
ggplot(data, aes(x=Senior, y=Exper, color = Sex)) + geom_point() + geom_smooth(method=lm,  linetype="dashed", se = FALSE, color="darkred", fill="blue") + labs(title = "Experience and Seniority", x = "Seniority", y = "Experience")
Inter.Senior.Exper.lm <- lm(Bsal ~ Senior + Exper + Senior:Exper, data=data)
summary(Inter.Senior.Exper.lm)
par(mfrow=c(2,2))
plot(Inter.Senior.Exper.lm)
```
# Model6A. Three-variable model with Bsal and Educ + Senior + Exper
```{r}
Educ.Senior.Exper.lm <- lm(Bsal ~ Educ + Senior + Exper, data=data)
summary(Educ.Senior.Exper.lm)
par(mfrow=c(2,2))
plot(Educ.Senior.Exper.lm)
```

# Model6B. Three-variable model with Bsal and poly(Educ, 2) + Senior + Exper
```{r}
PEduc.Senior.Exper.lm <- lm(Bsal ~ poly(Educ,2) + Senior + Exper, data=data)
summary(PEduc.Senior.Exper.lm)
par(mfrow=c(2,2))
plot(PEduc.Senior.Exper.lm)
```

# Model6C. Three-variable model with Bsal and Educ + Senior + Age
```{r}
Educ.Senior.Age.lm <- lm(Bsal ~ Educ + Senior + Age, data=data)
summary(Educ.Senior.Age.lm)
par(mfrow=c(2,2))
plot(Educ.Senior.Age.lm)
```
# Model7. Four-variable model with Bsal and Educ + Senior + Exper + Age
```{r}
Educ.Senior.Exper.Age.lm <- lm(Bsal ~ Educ + Senior + Exper + Age, data=data)
summary(Educ.Senior.Exper.Age.lm)
par(mfrow=c(2,2))
plot(Educ.Senior.Exper.Age.lm)
```

# Model8. Full model with Bsal and Educ + Senior + Exper + Age + Sex
```{r}
Full.lm <- lm(Bsal ~ Educ + Senior + Exper + Age + Sex, data=data)
summary(Full.lm)
par(mfrow=c(2,2))
plot(Full.lm)
```

#Model 9. Remove Exper from Model 8: model with Bsal and Educ + Senior + Age + Sex
```{r}
Educ.Senior.Age.Sex.lm <- update(Full.lm, .~.  - Exper)
summary(Educ.Senior.Age.Sex.lm)
par(mfrow=c(2,2))
plot(Educ.Senior.Age.Sex.lm)
```

#Model 10. Remove Age from Model 9: model with Bsal and Educ + Senior + Sex
```{r}
Educ.Senior.Sex.lm <- update(Educ.Senior.Age.Sex.lm, .~.  - Age)
summary(Educ.Senior.Sex.lm)
par(mfrow=c(1,4))
plot(Educ.Senior.Sex.lm)
plot(Educ.Senior.Sex.lm, 4)
coef(Educ.Senior.Sex.lm)
```

#Model 11. Remove Educ from Model 10: model with Bsal and Senior + Sex
```{r}
Senior.Sex.lm <- update(Educ.Senior.Sex.lm, .~.  - Educ)
summary(Senior.Sex.lm)
par(mfrow=c(2,2))
plot(Senior.Sex.lm)
```

#Model 12. Remove Senior from Model 11: model with Bsal and Sex
```{r}
Sex.lm <- update(Senior.Sex.lm, .~.  - Senior)
summary(Sex.lm)
par(mfrow=c(2,2))
plot(Sex.lm)
```

#AIC values
```{r}
AIC(Educ.lm, Senior.lm, Exper.lm, Age.lm, Educ.poly.lm, Senior.poly.lm, Exper.poly.lm, Educ.log.lm, Senior.log.lm, Exper.log.lm, Educ.Senior.lm, PEduc.Senior.lm, Educ.Exper.lm, Inter.Educ.Senior.lm, Inter.Educ.Exper.lm, Inter.Senior.Exper.lm, Educ.Senior.Exper.lm, Educ.Senior.Age.lm, Educ.Senior.Exper.Age.lm, Full.lm, Educ.Senior.Age.Sex.lm, Educ.Senior.Sex.lm, Senior.Sex.lm, Sex.lm)
```

# Model9. Exploration of interactions
```{r}
Full.Inter.lm <- lm(Bsal ~ Educ*Senior*Exper*Sex + I(Educ^2) + I(Senior^2) + I(Exper^2), data=data)
summary(Full.Inter.lm)
Full.Inter.lm.2 <- update(Full.Inter.lm, .~.  - Educ:Senior:Exper:Sex)
summary(Full.Inter.lm.2)
Full.Inter.lm.3 <- update(Full.Inter.lm.2, .~.  - Senior:Exper:Sex)
Full.Inter.lm.4 <- update(Full.Inter.lm.3, .~. -Educ:Exper:Sex)
Full.Inter.lm.5 <- update(Full.Inter.lm.4, .~. -Educ:Senior:Sex)
Full.Inter.lm.6 <- update(Full.Inter.lm.5, .~. -Educ:Senior:Exper)
Full.Inter.lm.7 <- update(Full.Inter.lm.6, .~. -Exper:Sex)
Full.Inter.lm.8 <- update(Full.Inter.lm.7, .~. -Senior:Sex)
Full.Inter.lm.9 <- update(Full.Inter.lm.8, .~. -Senior:Exper)
Full.Inter.lm.10 <- update(Full.Inter.lm.9, .~. -Educ:Sex)
Full.Inter.lm.11 <- update(Full.Inter.lm.10, .~. -Educ:Senior)
Full.Inter.lm.12 <- update(Full.Inter.lm.11, .~. -Educ:Exper)
Full.Inter.lm.13 <- update(Full.Inter.lm.12, .~. -I(Senior^2))
Full.Inter.lm.14 <- update(Full.Inter.lm.13, .~. -I(Educ^2))
Full.Inter.lm.15 <- update(Full.Inter.lm.14, .~. -I(Exper^2))
Full.Inter.lm.16 <- update(Full.Inter.lm.15, .~. -Exper)
summary(Full.Inter.lm.16)

AIC(Full.Inter.lm.16, Full.Inter.lm.12)
```

#Plot relationships with Bsal and Educ
```{r}
#Linear
par(mfrow=c(1,3))
plot(data$Educ, data$Bsal, main="Linear Model")
abline(Educ.lm, col="blue")

#non-linear
plot(data$Educ, data$Bsal,  main="Non-linear Model")
lines(data$Educ, predict(Educ.poly.lm), col="green") 

#transformed
plot(data$Educ, data$log_Bsal,  main="Transformed Model")
abline(Educ.log.lm, col="red")
```

#Plot relationships with Bsal and Senior
```{r}
#Linear
par(mfrow=c(1,3))
plot(data$Senior, data$Bsal, main="Linear Model")
abline(Senior.lm, col="blue")

#non-linear
plot(data$Senior, data$Bsal,  main="Non-linear Model")
lines(data$Senior, predict(Senior.poly.lm), col="green") 

#transformed
plot(data$Senior, data$log_Bsal,  main="Transformed Model")
abline(Senior.log.lm, col="red")
```

#Plot relationships with Bsal and Exper
```{r}
#Linear
par(mfrow=c(1,3))
plot(data$Exper, data$Bsal, main="Linear Model")
abline(Exper.lm, col="blue")

#non-linear
plot(data$Exper, data$Bsal,  main="Non-linear Model")
lines(data$Exper, predict(Exper.poly.lm), col="green") 

#transformed
plot(data$Exper, data$log_Bsal,  main="Transformed Model")
abline(Exper.log.lm, col="red")
```


#BACKWARD ELIMINATION

#Step 1. all variables
```{r}
data.lm.full <- lm(Bsal ~ Educ + Senior + Sex + Age + Exper, data = data)
summary(data.lm.full)
```

#Step 2. Exper removed
```{r}
data.lm.less.Exper <- update(data.lm.full, .~. - Exper, data = data)
summary(data.lm.less.Exper)
```

#Step 3. Age removed - last step, all p-values < 0.01
```{r}
Sex.Senior.Educ.lm <- update(data.lm.less.Exper, .~. - Age, data = data)
summary(Sex.Senior.Educ.lm)
#residual analysis
par(mfrow = c(2,2))
plot(Sex.Senior.Educ.lm)
plot(Sex.Senior.Educ.lm, 4) #no influential points
```


# SUBSETS

```{r}
install.packages(("caret"))
library(caret)
# Set seed for reproducibility
set.seed(123)
# Set up repeated k-fold cross-validation
train.control <- trainControl(method = "cv", number = 10)
# Train the model
step.model <- train(Bsal ~., data = data,
                    method = "leapBackward", 
                    tuneGrid = data.frame(nvmax = 1:5),
                    trControl = train.control
                    )
summary(step.model)
# Model accuracy
step.model$results
#how many variables to use
plot(step.model$results[,"nvmax"], step.model$results[,"Rsquared"], type = "l")
plot(step.model$results[,"nvmax"], step.model$results[,"MAE"], type = "l")
plot(step.model$results[,"nvmax"], step.model$results[,"Rsquared"], type = "p")
```

#Best tuning values
```{r}
# best tuning values
step.model$bestTune
```

```{r}
# Summary of the model
summary(step.model$finalModel)
```

```{r}
coef(step.model$finalModel, 3)
```

```{r}
anova(Educ.Senior.lm, Sex.Senior.Educ.lm)
```
